Avoiding bias in the model fitting of correlated interferometric data

### 2021-01-27 article.tex
almost ready

\documentclass[a4paper,fleqn,usenatbib]{mnras}
\usepackage[T1]{fontenc}

\usepackage{amsmath,amsbsy}
\usepackage{graphicx}
%\usepackage{natbib}
\bibliographystyle{mnras}
\usepackage{hyperref}

@@

\title{Avoiding bias in the model fitting of correlated interferometric data}
\author{R\'egis Lachaume}
\maketitle

\begin{abstract}
I show how Peelle's pertinent puzzle can be avoided in the framework of the least squares minimisation.
\end{abstract}

@@

Unfortunately, it has been shown that ignoring uncertainties may lead to significant errors in model paremeters as \citet{LAC19} showed with stellar diameters with PIONIER\footnote{Precision Integrated-Optics Near-infrared Imaging ExpeRiment} \citep{PIONIER} data at the VLTI. Also \citet{KAM20} established that correlations were necessary to achieve a higher contrast ratio in companion detection using GRAVITY \citep{GRAVITY} at the VLTI.

@@

Let's consider a set of measurements of a single normalised quantity, like the visibility amplitude, given by a vector $\vdata = \tr{(\data_1, \cdots, \data_n)}$. It is derived from an uncalibrated quantity like the fringe contrast, $\vraw = \tr{(\raw_1, \cdots, \raw_n)}$ and a normalisation factor, like the cotransfer function, $\vcot = \tr{(\cot_1, \cdots, \cot_n)}$ by 
\begin{equation}
    \vdata = \vcot\hadam\vraw,
\end{equation}
where $\hadam$ denotes the Hadamard (elementwise) product of vectors. If $\datamean$, $\cotmean$, and $\rawmean$ the true, but unknown, values of these quantities, the error vector on $\vdata$  
\begin{align}
    \verror    &= \vdata - \datamean, \\
\intertext{can be written as a sum of measurement and normalisation errors if one ignores a second-order term:}
    \verror    &= \vabserror + \datamean \vrelerror.\\
\intertext{These errors are given by}
    \vabserror &= \cotmean (\vraw - \rawmean), \\
    \vrelerror &= \frac1\cotmean (\vcot-\cotmean).
\end{align}

@@

Since $\expect{\error_i} = 0$ and $\expect{(\error_i-\error_j)^2} = 2\devmean^2$, the expectancy
\begin{equation}
    \expect{\ppp{\mod}} \approx \datamean 
            \left[ 1 -  
             \left(1-\frac1n\right)
             \left(2 + (n-2)\corr\right)
              \reldev^2 \right]
\end{equation}
is biased.  If the data are not correlated ($\corr = 0$), the bias is small
($\reldev^2$ to $2\reldev^2$) but it becomes large for correlated data if the number of points is large ($\sim n\reldev^2$ for fully correlated data) as \citet{DAG94} already noted. Figure~\ref{fig:uncorr-peelle} shows a simulation of the bias as a function of the normalisation uncertainty $\reldev$ for various data sizes ($n = 2$ to 100). 
\begin{figure}
\centering
\includegraphics[width=\linewidth]{pdf/uncorrelated-peelle.pdf}
\caption{Fit $\datappp$  to unresolved visibilities ($\data = 1$), as a function of the relative uncertainty on the calibration $\reldev$ and the number of measurements $n$. $2/n\times10^5$ simulations were made and averaged, assuming that $\abserror$ and $\relerror$ follow normal distributions. \textit{Top:} fully correlated normalisation like in original Peelle's puzzle ($\absdev=0.02$ and $\corr = 1$). \textit{Bottom:} normalisation error without correlation ($\absdev=0.02$ and $\corr = 0$).}
\label{fig:uncorr-peelle}
\end{figure}

@@

\begin{align}
    \vmod  &= \msens\vparam,\\
\intertext{where $\vparam$ are the unknown parameters and $\msens$ is the known sensitivity matrix. Typically, $\sens_{ik}=f_k(u_i, v_i)$ for a linear model and $\sens_{ik} = \partial f/\partial p_k(u_i, v_i)$ for a non-linear model approximated by a linear one close to a solution. $(u, v)$ is the reduced baseline.  The true values $\vdatamean$ are are impacted by errors so that the data are} 
    \vdata &= \vdatamean + \verror\\
\intertext{with the error term $\verror$ again expressed as the sum of a measurement error and a normalisation one:}
    \verror &=  \vabserror + \vrelerror \hadam \vdatamean.
\end{align}
The measurement errors $\vec\abserror$ and normalisation errors $\vec\relerror$ following multivariate distributions of mean zero and covariance matrices $\vabscov$ and $\vrelcov$ respectively. Given the covariance matrix $\vcov$ of this model, the least squares estimate is
\begin{align}
   \vparam &= (\tr\msens \vcov^{-1} \msens)^{-1} (\tr\msens \vcov^{-1} \vdata)
\end{align}
I investigate four ways to determine the covariance matrix
\begin{enumerate}
    \item Ignoring the correlations in the normalisation using $\vcov_0 = \vabscov + (\vdata\outer\vdata)\hadam(\vrelcov\hadam\vec I)$. Let $\vdataflat = \msens(\tr\msens\vcov_0^{-1}\msens)^{-1}\tr\msens\vcov_0^{-1}\vdata$ the resulting model of the data.\label{item:nocorr} 
    \item Using the na\"\i{}ve estimate $\ppp\vcov = \vabscov + (\vdata\outer\vdata)\hadam\vrelcov$ which is known to lead to Peelle's pertinent puzzle in the trivial case of a constant model.\label{item:ppp}
    \item Using the data model of the fit without the normalisation error: $\vcov_1 = \vabscov + (\vdataflat\outer\vdataflat)\hadam\vrelcov$. This is the generalisation of the two-variables approach by \citet{NEU14}. The resulting least squares model is $\vdatarec{1} = \msens(\tr\msens\vcov_1^{-1}\msens)^{-1}\tr\msens\vcov_1^{-1}\vdata$.\label{item:fit}
    \item Recursively fitting the data by updating the data model in the covariance matrix. I derive $\vdatarec{k} = \msens(\tr\msens\vcov_k^{-1}\msens)^{-1}\tr\msens\vcov_k^{-1}\vdata$ using $\vcov_{k} = \vabscov + (\vdatarec{k-1}\outer\vdatarec{k-1})\hadam\vrelcov$, starting with the estimate $\vdatarec{1}$ ($k = 2$).\label{item:recfit}
\end{enumerate}

Figure~\ref{fig:fitexample} shows the example of a typical centrosymmetric quadratic fit to under-resolved data obtained at a four-telescope facility in medium spectral resolution mode.  Given the dimensionless variable $x$ proportional to the projected baseline length ($x \propto \sqrt{u^2+v^2}$), a model fit $\mod = a - bx^2$ is performed to simulated data following $\data \approx 1 - x^2$, including a (correlated) systematic normalisation error $\relerror$ (3\%) and independent statistical errors $\abserror$ (2\%).  As expected, the use of data $\vdata$ in the correlation matrix (method \ref{item:ppp}) leads to grossly underestimated data values, in the very same way as in the classical Peelle case described in Sects.~\ref{sec:ppp}~\&~\ref{sec:single}.  Other methods, including ignoring correlations, yield reasonable fits.

Figure~\ref{fig:fitanalysis} sums up the behaviour of the same fit performed a large number of times on different simulated data sets, each following $\data \approx 1-x^2$.  For each correlation matrix prescription, it displays the dispersion of the reduced chi squared, the model parameters $a$ and $b$, and the difference between modelled value and true value. It also reports the uncertainty on model parameters reported by the least squares fit in comparison to the scatter of the distribution of the values. Both methods estimating the correlation  matrix from modelled data (\ref{item:fit} \& \ref{item:recfit}) are equivalent in terms of the absence of bias, dispersion of these quantities, and correct prediction of the uncertainty on model parameters.  While the model fits ignoring correlations (\ref{item:nocorr}) are not biased either, they display a higher dispersion of model parameters and grossly underestimate the uncertainty on model parameters.  The correlation matrix calculated from data (\ref{item:ppp}) is, as expected, strongly biased.

\section{Conclusion}
\begin{enumerate}
\item Peelle's pertinent puzzle primarily arises from an incorrect description of the relative uncertainties.  It will occur in a fully linear problem and, to some extent, without correlations.
\item For uncorrelated data, the relative bias is of the order of the square of the relative uncertainty as long as errors remain small. It increases twofold when the number of data points becomes large.
\item For correlated data, the relative bias is larger by a factor of the order of the number of correlated points as long as the resulting error remain small.
\item In least-squares model fits, the problem disappears if the uncertainties, or the covariance matrix, are updated with the model values instead of the measurement values.  
\end{enumerate}

>>>>

\documentclass[a4paper,fleqn,usenatbib]{mnras}
\pdfoutput=1

\usepackage{amsmath,amsbsy}
\usepackage{graphicx}
\bibliographystyle{mnras}
\usepackage{hyperref}

@@

\title{Avoiding bias in the model fitting of correlated interferometric data}
\author{R\'egis Lachaume$^{1,2}$\thanks{regis.lachaume@gmail.com}
\\
$^{1}$Instituto de Astronom\'\i{}a, Facultad de F\'\i{}sica, Pontificia Universidad Cat\'olica de Chile, casilla 306, Santiago 22, Chile\\
$^{2}$Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany
}
\maketitle

\begin{abstract}
In optical and infrared long-baseline interferometry, data often display significant correlated errors because of uncertain multiplicative factors such as the instrumental transfer function or the pixel-to-visibility matrix.  In the context of model fitting, this situation often leads to a significant bias, which in the gravest cases can show a fit passing outide of all error bars. It is known in nuclear physics as Peelle's Pertinent Puzzle.  I show how this arises in the context of interferometry and how severe it can be in a realistic setting.  I then give a conceptually simple and computationnally cheap way to avoid the issue: model the data without covariances, estimate the covariance matrix by error propagation using the modelled data instead of the actual data, and, finally, perform the model fitting using the covariance matrix.
\end{abstract}

\begin{keywords}
techniques: interferometric --- 
methods: statistical
\end{keywords}

@@

Unfortunately, it has been shown that ignoring uncertainties may lead to significant errors in model paremeters as \citet{LAC19} evidenced with stellar diameters using PIONIER\footnote{Precision Integrated-Optics Near-infrared Imaging ExpeRiment} \citep{PIONIER} data at the VLTI. Also \citet{KAM20} established that correlations were necessary to achieve a higher contrast ratio in companion detection using GRAVITY \citep{GRAVITY} at the VLTI.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{pdf/uncorrelated-peelle.pdf}
\caption{Fit $\datappp$  to unresolved visibilities ($\data = 1$), as a function of the relative uncertainty on the calibration $\reldev$ and the number of measurements $n$. $2/n\times10^5$ simulations were made and averaged, assuming that $\abserror$ and $\relerror$ follow normal distributions. \textit{Top:} fully correlated normalisation like in original Peelle's puzzle ($\absdev=0.02$ and $\corr = 1$). \textit{Bottom:} normalisation error without correlation ($\absdev=0.02$ and $\corr = 0$).}
\label{fig:uncorr-peelle}
\end{figure}

Let's consider a set of measurements of a single normalised quantity, like the visibility amplitude, given by a column vector $\vdata = \tr{(\data_1, \cdots, \data_n)}$. It is derived from an uncalibrated quantity like the fringe contrast, $\vraw = \tr{(\raw_1, \cdots, \raw_n)}$ and a normalisation factor, like the cotransfer function, $\vcot = \tr{(\cot_1, \cdots, \cot_n)}$ by 
\begin{equation}
    \vdata = \vcot\hadam\vraw,
\end{equation}
where $\hadam$ denotes the Hadamard (elementwise) product of vectors. If $\datamean$, $\cotmean$, and $\rawmean$ the true, but unknown, values of these quantities, the error vector on $\vdata$  
\begin{align}
    \verror    &= \vdata - \datamean, \\
\intertext{can be written as a sum of measurement and normalisation errors if one ignores a second-order term:}
    \verror    &= \vabserror + \datamean \vrelerror.\\
\intertext{These errors are given by}
    \vabserror &= \cotmean (\vraw - \rawmean), \\
    \vrelerror &= \frac1\cotmean (\vcot-\cotmean).
\end{align}

@@

Since $\expect{\error_i} = 0$ and $\expect{(\error_i-\error_j)^2} = 2\devmean^2$, the expectancy
\begin{equation}
    \expect{\ppp{\mod}} \approx \datamean 
            \left[ 1 -  
             \left(1-\frac1n\right)
             \left(2 + (n-2)\corr\right)
              \reldev^2 \right]
\end{equation}
is biased.  If the data are not correlated ($\corr = 0$), the bias is small
($\reldev^2$ to $2\reldev^2$) but it becomes large for correlated data if the number of points is large ($\sim n\reldev^2$ for fully correlated data) as \citet{DAG94} already noted. Figure~\ref{fig:uncorr-peelle} shows a simulation of the bias as a function of the normalisation uncertainty $\reldev$ for various data sizes ($n = 2$ to 100).

@@

\begin{align}
    \vmod  &= \msens\vparam,\\
\intertext{where $\vparam$ are the unknown parameters and $\msens$ is the known sensitivity matrix. Typically, $\sens_{ik}=f_k(u_i, v_i)$ for a linear model and $\sens_{ik} = \partial f/\partial p_k(u_i, v_i)$ for a non-linear model approximated by a linear one close to a solution. $(u, v)$ is the reduced baseline.  The true values $\vdatamean$ are are impacted by errors so that the data are} 
    \vdata &= \vdatamean + \verror\\
\intertext{with the error term $\verror$ again expressed as the sum of a measurement error and a normalisation one:}
    \verror &=  \vabserror + \vrelerror \hadam \vdatamean.
\end{align}
The measurement errors $\vec\abserror$ and normalisation errors $\vec\relerror$ following multivariate distributions of mean zero, with covariance matrices $\vabscov$ and $\vrelcov$ respectively. Given the covariance matrix $\vcov$ of this model, the least squares estimate is
\begin{align}
   \vparam &= (\tr\msens \vcov^{-1} \msens)^{-1} (\tr\msens \vcov^{-1} \vdata)
\end{align}
I investigate four ways to determine the covariance matrix
\begin{enumerate}
    \item Ignoring the correlations in the normalisation using $\vcov_0 = \vabscov + (\vdata\outer\vdata)\hadam(\vrelcov\hadam\vec I)$. Let $\vdataflat = \msens(\tr\msens\vcov_0^{-1}\msens)^{-1}\tr\msens\vcov_0^{-1}\vdata$ the resulting model of the data.\label{item:nocorr} 
    \item Using the na\"\i{}ve estimate $\ppp\vcov = \vabscov + (\vdata\outer\vdata)\hadam\vrelcov$ which is known to lead to Peelle's pertinent puzzle in the trivial case of a constant model.\label{item:ppp}
    \item Using the data model of the fit without the normalisation error: $\vcov_1 = \vabscov + (\vdataflat\outer\vdataflat)\hadam\vrelcov$. This is the generalisation of the two-variables approach by \citet{NEU14}. The resulting least squares model is $\vdatarec{1} = \msens(\tr\msens\vcov_1^{-1}\msens)^{-1}\tr\msens\vcov_1^{-1}\vdata$.\label{item:fit}
    \item Recursively fitting the data by updating the data model in the covariance matrix. I derive $\vdatarec{k} = \msens(\tr\msens\vcov_k^{-1}\msens)^{-1}\tr\msens\vcov_k^{-1}\vdata$ using $\vcov_{k} = \vabscov + (\vdatarec{k-1}\outer\vdatarec{k-1})\hadam\vrelcov$, starting with the estimate $\vdatarec{1}$ ($k = 2$).\label{item:recfit}
\end{enumerate}

In order to compare these covariance matrix prescriptions, I will use the typical example of an under-resolved centrosymmetric source observed at a four-telescope facility in high spectral resolution. In the under-resolved case all models---Gaussian, uniform disc, or limb-darkened disc---are equivalent \citep{LAC03}--- so I will use instead a linear least squares fit $\mod = a - bx^2$ to $\data \approx 1 - x^2$ where $x$ is dimensionless variable proportional to the projected baseline length $\sqrt{u^2+v^2}$. This fit corresponds to the second-order Taylor development of any of the aforementionned models.  Figure~\ref{fig:fitexample} shows the example of such a fit performed for each covariance matrix prescription.  Data has been simulated using $\data = (1-x^2)(1 + \relerror) + \abserror$ where $\relerror$ is a fully correlated normalisation error (3\%) and $\abserror$ are uncorrelated statistical errors (2\%). As expected, the use of data $\vdata$ in the correlation matrix (method \ref{item:ppp}) leads to grossly underestimated data values, in the very same way as in the classical Peelle case described in Sects.~\ref{sec:ppp}~\&~\ref{sec:single}.  Other methods, including ignoring correlations, yield reasonable fits.

Figure~\ref{fig:fitanalysis} sums up the behaviour of the same fit performed a large number of times on different simulated data sets, each following $\data \approx 1-x^2$.  For each correlation matrix prescription, it displays the dispersion of the reduced chi squared, the model parameters $a$ and $b$, and the difference between modelled value and true value. It also reports the uncertainty on model parameters reported by the least squares fit in comparison to the scatter of the distribution of the values. While the model fits ignoring correlations \ref{item:nocorr} does not show any bias, they display a higher dispersion of model parameters and grossly underestimate the uncertainty on model parameters.  The correlation matrix calculated from data \ref{item:ppp} is, as expected, strongly biased. Both methods estimating the correlation  matrix from modelled data, \ref{item:fit} \& \ref{item:recfit}, are equivalent in terms of the absence of bias, dispersion of these quantities, and correct prediction of the uncertainty on model parameters.

Given that fitting recursively the covariance matrix doesn't yield additional benefits, I would suggest to use method \ref{item:fit}.

\section{Conclusion}

The standard, but incorrect, covariance propagation in any ``normalisation'' setting leads to bias in the model parameters of a least-squares fit taking correlations into account.  Some bias will even occur without correlations, but the effect is strongest when a large set of correlated data is modelled.  This is precisely the case in optical and infrared long-baseline interferometry, where the calibration of spectrally dispersed fringes easily yields $10^2$ to $10^3$ correlated data points.

While solutions exist that are either numerically expensive or require some care to be implemented \citep{BUR11,BEC12,NIS14}, I have shown with a simple example that there is an easy and cheap way to solve the issue. First a uncorrelated fit is performed to estimate the true values corresponding to the data.  Secondly, these estimates are used to determine the covariance matrix by error propagation. At last, this covariance matrix is used to perform a model-fit.
